# Organizing and Displaying Data

## Introduction

*R* and RStudio are two great tools for creating graphs. Although I think SPSS scores high on the graph creation category, it falls short in many aspects. Arguably, its main shortcoming has to do the lack of flexibility. This is not the case with *R*.

**Requirements**

-   Install *R* and RStudio[^ch-graphing-data-1] - Section \@ref(package-download-installation)

[^ch-graphing-data-1]: Alternatively, you can use RStudio Cloud - www.rstudio.cloud

Below, I will show how to create beautiful graphs with *R* and RStudio. In addition, I will provide the script that can be used to create the same graphs in SPSS.

## Line Graphs

**Packages**

Install and load the following packages before proceeding. To accomplish this, copy, paste, and run the following in **Console**.

`install.packages("ggplot2") # will install the ggplot2 package`

`library("ggplot2") # will load the ggplot2 package`

I will demonstrate below how to create **single line** graphs and **multiple lines** graph using RStudio and SPSS.

To follow along, open RStudio, then create a new *R* script[^ch-graphing-data-2]. Then, copy and paste each chunk of code into the newly created script.

[^ch-graphing-data-2]: This is optional; you can run the script directly in the `Console` panel in RStudio

Finally, run *separately* each step (chunks of code) to create the graphs. We will start with Step 1, then run Step 2 to create the graph shown in Figure \@ref(fig:simple-line-1). Notice that this graph does not have the lines (or titles), yet.

### Simple line graph

**Step 1**

```{r}
# Step 1: Create two variables (x, y) and assign numbers to each
x <- c(1:5)
y <- c(12,18,22,20,28)
```

**Step 2**

```{r simple-line-1, fig.cap="Simple graph line without title nor lines"}
# Step 2: Create a simple plot
plot(x, y)
```

**Step 3**

```{r simple-line-2, fig.cap="Simple graph with titles but no lines"}
# Step 3: Let's add labels for x and y and a title for the plot
plot(x, y, xlab = "x var title", ylab = "y var title", main = "Title")
```

**Step 4**

```{r simple-line-3, fig.cap="Simple graph line with titles and lines"}
# Step 4: Let's draw a line
plot(x, y, xlab = "x var title", ylab = "y var title", main = "Title", 
     type = "o")
```

**Step 5**

```{r simple-line-4, fig.cap="Simple graph line with titles, different line density, and color"}
# Step 5: Let's change the density and color of the line and dots
plot(x, y, xlab = "x var title", ylab = "y var title", main = "Title", 
     type = "o", pch = 20, lwd = 2, col = "blue")
```

### Multiple lines graph

The R script below was adapted from: <http://bit.ly/34fDitr>

To accomplish this task, you will need to use three other functions in R: points() and lines()

**What you need to know**

Three groups were tested 5 times (once every week for a period of 5 weeks) on BMI (body mass index). For a period of five weeks, two groups engaged in a physical fitness program whereas G1 did not exercise at all. Using this information, plot BMI changes over time for the three groups.

    Group 1 (G1) = No exercise (control group)
    Group 2 (G2) = Traditional fitness program
    Group 3 (G3) = New fitness program

**Step 1**

```{r}
# define 3 data sets
time <- c(1,2,3,4,5) # or simply c(1:5)
y1 <- c(24,24,26,23,25)      # Group 1
y2 <- c(28, 25, 26, 25, 24)  # Group 2
y3 <- c(30,30, 28, 26, 25)   # Group 3
```

**Step 2**

```{r multi-line-1, fig.cap="Multi line graph - first line"}
# plot the first curve by calling plot() function
# First curve is plotted
plot(time, y1, type="o", col="blue", pch="o", lty=1, ylim=c(0,50))
```

**Step 3**

```{r multi-line-2, fig.cap="Multi line graph - second line added"}
# Add second curve to the same plot by calling points() and lines()
# Use symbol '*' for points.
plot(time, y1, type="o", col="blue", pch="o", lty=1, ylim=c(0,50))
points(time, y2, col="red", pch='*')
lines(time, y2, col="red",lty=2)
```

**Step 4**

```{r multi-line-3, fig.cap="Multi line graph - third line added"}
# Add Third curve to the same plot by calling points() and lines()
# Use symbol '+' for points.
plot(time, y1, type="o", col="blue", pch="o", lty=1, ylim=c(0,50))
points(time, y2, col="red", pch='*')
lines(time, y2, col="red",lty=2)
points(time, y3, col="dark red",pch="+")
lines(time, y3, col="dark red", lty=3)
```

{{% alert note %}} After running the script above, you will notice an empty space on the top-half of the plot. So, change the following: (1) Change the `ylim` argument to `20, 35`. This will improve the look of the plot. (2) In addition, add `labels` to `y` and `x` and a `main title` to the plot. Refer to Figure \@ref(fig:simple-line-2) for more details. {{% /alert %}}

**Step 5**

```{r multi-line-4, fig.cap="Multi line graph - remove space and add label"}
# Add Third curve to the same plot by calling points() and lines()
# Use symbol '+' for points.
plot(time, y1, xlab = "x var title", ylab = "y var title", main = "Title", type="o", col="blue", pch="o", lty=1, ylim=c(20,35))
points(time, y2, col="red", pch='*')
lines(time, y2, col="red",lty=2)
points(time, y3, col="dark red",pch="+")
lines(time, y3, col="dark red", lty=3)
```

### Multiple lines graph (with legends)

Below is the R script to create a plot with multiple lines and a legend. To do this, create a new script, then paste the script below into RStudio and run the script the same way it was done previously.

**Step 1**

```{r}
# define 3 data sets
xtime <- c(1,2,3,4,5) # or simply c(1:5)
y1 <- c(24,24,26,23,25)     # Group 1
y2 <- c(28,25,26,25,24)     # Group 2
y3 <- c(30,30,28,26,25)     # Group 3
```

**Step 2**

```{r}
# plot the first curve by calling plot() function
# First curve is plotted
plot(xtime, y1, type="o", col="blue", pch="o", lty=1, ylim=c(20,40), ylab="y" )
```

**Step 3**

```{r}
# Add second curve to the same plot by calling points() and lines()
# Use symbol '*' for points.
plot(xtime, y1, type="o", col="blue", pch="o", lty=1, ylim=c(20,40), ylab="y" )
points(xtime, y2, col="red", pch="*")
lines(xtime, y2, col="red",lty=2)
```

**Step 4**

```{r}
# Add Third curve to the same plot by calling points() and lines()
# Use the symbol '+' for points.
plot(xtime, y1, type="o", col="blue", pch="o", lty=1, ylim=c(20,40), ylab="y" )
points(xtime, y2, col="red", pch="*")
lines(xtime, y2, col="red",lty=2)
points(xtime, y3, col="dark red",pch="+")
lines(xtime, y3, col="dark red", lty=3)
```

**Step 5**

```{r}
# Adding a legend inside the box at the location (1,37) in graph coordinates.
# Note that the order of plots are maintained in the vectors of attributes.
plot(xtime, y1, type="o", col="blue", pch="o", lty=1, ylim=c(20,40), ylab="y" )
points(xtime, y2, col="red", pch="*")
lines(xtime, y2, col="red",lty=2)
points(xtime, y3, col="dark red",pch="+")
lines(xtime, y3, col="dark red", lty=3)
legend(1,37,legend=c("y1","y2","y3"), col=c("blue","red","black"),
       pch=c("o","*","+"),lty=c(1,2,3), ncol=1)
```

### Line Graphs in SPSS

    GGRAPH
      /GRAPHDATASET NAME="graphdataset" VARIABLES=x MEAN(y)[name="MEAN_y"] MISSING=LISTWISE 
        REPORTMISSING=NO
      /GRAPHSPEC SOURCE=INLINE.
    BEGIN GPL
      SOURCE: s=userSource(id("graphdataset"))
      DATA: x=col(source(s), name("x"), unit.category())
      DATA: MEAN_y=col(source(s), name("MEAN_y"))
      GUIDE: axis(dim(1), label("x"))
      GUIDE: axis(dim(2), label("Mean y"))
      GUIDE: text.title(label("Simple Line Mean of y by x"))
      SCALE: linear(dim(2), include(0))
      ELEMENT: line(position(x*MEAN_y), missing.wings())
    END GPL.

## Boxplots

## Bar Graphs

## Scatter Plots

## Normality Tests with Plots

Several tests (t-test, ANOVA, Pearson Coefficient) assume the independent variable is normally distributed. This is especially important when the sample size is small (less than 30) and not randomly drawn from the population. Typically, when the sample size is 30 or more and a random technique is used, violating the normality assumption is less of a problem. This is because of the central limit theorem, discussed in section ??.

Below, I will discuss four methods to assess normality. These are 1) histogram, 2) QQ plots, 3) skewness/kurtosis, and 4) Shapiro\_Wilk test.

### Histograms

A histogram can help identifying patterns in a data set. For instance, one can visually identify potential deviation of normality (skewness and kurtosis) with a histogram, which is simply a frequency distribution of a data set. To create a histogram, we plot the frequency of observations against data points allocated into specific intervals.

For this particular example, we will be using a data set called `women` that comes with *R*. The data gives the average heights and weights for American women aged 30--39. Visit this link to learn more about the data set: <https://bit.ly/352W2he>

### Load the data

Calling the following function will load the data in **Environment**. Now you are ready to create the histogram.

```{r}
data("women")
```

### Summary

Before creating the histogram, let's use a function to summarize the data and learn a bit more about the data set. Notice that the built-in function in *R* does not calculate the standard deviation as part of the summary. There are other functions that provide a more complete list of measures of central tendency and dispersion. For now, we can call another function to calculate the standard deviation of each of the variables below.

```{r}
summary(women)  # summary of the data set
sd(women$height) # calculates the standard deviation for height
sd(women$weight) # calculates the standard deviation for weight
```
